---
title: "R Notebook"
output: html_document
---

## Overview

This serves to document how I put together my dataset for the COVID-19 county classifier I'm trying to make (still in progress). 

```{r}
library(tidyverse)
library(fuzzyjoin)
```

Merge raw county level data into a single dataframe, containing:
- The county name, state, and state code
- The total population size
- Health-related attributes of the county (diabetes prevalence, air quality, etc.)
- Percent of the population over 65
```{r}
# read local data into memory
county_health <- read_csv("../data/raw/health_us_county_part1.csv")
county_health_2 <-read_csv("../data/raw/health_us_county_part2.csv")
state_abv <- read_csv("../data/raw/state_abv.csv")
population_data <- read_tsv("../data/raw/county_pop_estimates_2019.csv")
perc_over_65 <- read_csv("../data/raw/perc_over_65.csv")

# remove leading period that appeared when I converted this to a csv
population_data$`Geographic Area` <- gsub('^\\.|\\.$', '', population_data$`Geographic Area`)

# Merge local data into a single dataframe, one row per county
county_health <- county_health %>%
  left_join(county_health_2)
population_data <- population_data %>%
  separate(`Geographic Area`, c("County", "State"), c(", "))
county_health <- county_health %>% 
  separate(Geography, c("County", "Code"), c(", ")) %>%
  full_join(state_abv) %>%
  full_join(population_data, by = c("County" = "County", "State" = "State"))
# this strips the string " County" from entries in the County column
county_health$County <- sub("\\s+(?:County)$", "", county_health$County)
perc_over_65$County <- sub("\\s+(?:County)$", "", perc_over_65$County)
county_health <- county_health %>% 
  left_join(perc_over_65) %>%
  rename(`Percent Over 65` = "Percent")
county_health
```

Merge the county-level health data with COVID-19 case counts and death counts
```{r}
# grab COVID cases & deaths (by county) dataset from the internet
covid_data <- read_csv("https://github.com/nytimes/covid-19-data/raw/master/us-counties.csv")
covid_data <- covid_data %>% filter(date == "2020-11-20")
# merge covid case/death counts per county with county health data
joined <- county_health %>% 
   filter(Year == 2018) %>% 
   left_join(covid_data, by = c("County" = "county", "State" = "state"))
```

Drop unneeded columns and write the final dataset to disk
```{r}
final <- select(joined, -c("Year", `ID Geography`, "date"))
final %>% write_csv("../data/health_metrics_and_covid_cases_by_county.csv")
```

