---
title: "R Notebook"
author: "Alexis Hill"
output: html_notebook
---

## Overview

This code reads in the data set documented by `county_atributes_and_covid_severity.Rmd` and does a PCA (on the state level) to get a better feel for the data. 

### Load libraries
```{r}
#install_github("vqv/ggbiplot")
library(tidyverse)
library(devtools)
library(ggbiplot)
```

### Read in data
```{r}
county_health_covid <- read_csv("../data/health_metrics_and_covid_cases_by_county.csv")
# create two new "per capita" columns for deaths and cases, respectively
county_health_covid <- county_health_covid %>% 
  mutate(`Deaths per capita` = deaths / `Population`, 
         `Cases per capita` = cases / `Population`)
state_choice <- c("Florida") # select a state for the analysis
```

### Some misc exploratory plotting
```{r}
county_health_covid %>% filter(State == state_choice) %>% ggplot() +
  aes(`Deaths per capita`) %>% 
  geom_histogram()

county_health_covid %>%
  filter(State == state_choice) %>%
  ggplot() +
  aes(Percent_Over_65, `Deaths per capita`) +
  geom_point()
```

### Grab the state specific subset of the data, and create a new grouping `Severity` based on the deaths per capita in each county
```{r}
subset_by_state <- county_health_covid %>% filter(State == state_choice)
std_dev_deaths_cap <- sd(subset_by_state$`Deaths per capita`)
mean_deaths_cap <- mean(subset_by_state$`Deaths per capita`)

# Creates a Severity label (low, med, high) based on the deaths per capita in each county
subset_by_state <- subset_by_state %>% mutate(Severity = case_when(`Deaths per capita` <= (mean_deaths_cap - std_dev_deaths_cap) ~ "low",
                                                                   `Deaths per capita` > (mean_deaths_cap - std_dev_deaths_cap) &
                                                                   `Deaths per capita` <= (mean_deaths_cap + std_dev_deaths_cap) ~ "med",
                                                                   `Deaths per capita` > (mean_deaths_cap + std_dev_deaths_cap) ~ "high"))

# Same as above, but will split the "med" group into "med-low" and "med-high" groups
# subset_by_state <- subset_by_state %>% mutate(Severity = case_when(`Deaths per capita` <= (mean_deaths_cap - std_dev_deaths_cap) ~ "low",
#                                                                    `Deaths per capita` > (mean_deaths_cap - std_dev_deaths_cap) &
#                                                                    `Deaths per capita` <= mean_deaths_cap ~ "med low",
#                                                                    `Deaths per capita` > mean_deaths_cap &
#                                                                    `Deaths per capita` <= (mean_deaths_cap + std_dev_deaths_cap) ~ "med high",
#                                                                    `Deaths per capita` > (mean_deaths_cap + std_dev_deaths_cap) ~ "high"))
```

### For the actual PCA, create another subset with only numeric values, and drop unneeded variables (note: we'll use `Severity` later to color our PCA plot)
```{r}
subset_by_state_numeric <- subset_by_state[,sapply(subset_by_state, is.numeric)] 
subset_by_state_numeric <- select(subset_by_state_numeric, -c("Population", "cases", "deaths", `Deaths per capita`, `Cases per capita`))
# scale data and replace NAs with 0
subset_by_state_numeric <- scale(subset_by_state_numeric)
subset_by_state_numeric[is.na(subset_by_state_numeric)] <- 0
```

### Finally, run the PCA and plot the first three PCs
```{r}
#pairs(subset_by_state_numeric)
state.pca <- prcomp(subset_by_state_numeric)
summary(state.pca)
ggbiplot(state.pca, labels = subset_by_state$County, ellipse=TRUE, groups=subset_by_state$Severity)
ggbiplot(state.pca, choices = c(1,3), labels = subset_by_state$County, ellipse=TRUE, groups=subset_by_state$Severity)
```